On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-22 manuscripts/PNAS/chazthm_pnas.tex
last sml edits

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{B}lind source separation is a classical problem in signal processing \cite{sato1975method}.
In one common modern formulation, each of $N$ observed $n$-dimensional signals is a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary'' of size $m$, typically with $m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models).
Approximating solutions to this sparsity-constrained inverse problem have provided insight into the structure of many signal classes lacking domain-specific formal models (e.g., in vision \cite{wang2015sparse}).  In particular, it has been shown that response properties of simple-cell neurons in mammalian visual cortex emerge from optimizing a dictionary to represent small patches of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major advance in computational neuroscience. A curious aspect of this finding is that the latent waveforms (e.g., ``Gabor'' wavelets) estimated from data appear to be canonical \cite{donoho2001can};
i.e., they are found in learned dictionaries independent of algorithm or natural image training set.

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if all dictionaries consistent with training data do not differ appreciably in their ability to sparsely encode new samples. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain this uniqueness phenomenon more broadly; in particular, a universal guarantee in the context of noise has yet to emerge.

@@

\begin{problem}[Sparse linear coding]\label{InverseProblem}
Find a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ such that $\|\mathbf{z}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \eta$ for all $i$.
\end{problem}

@@

\vspace{-.15 cm}

@@

\begin{lemma}\label{spanIntersectionLemma}
If $f: U \to V$ is an injective function, then $f\left(\cap_{i=1}^\ell U_i \right) =  \cap_{i=1}^\ell f\left(U_i\right)$ for any $U_1, \ldots, U_\ell \subseteq U$.
%In particular, if for some $E \in 2^{[m]}$ the map $M \in \mathbb{R}^{n \times m}$ is injective on $\cup_{S \in \mathcal{H}} \text{span}\{e_i\}_{i \in S}$ then $\text{span}\{ M_{\cap_{S \in \mathcal{H}} S} \} = \cap_{S \in \mathcal{H}} \text{span}\{M_S\}$.
\end{lemma}
\begin{proof}
By induction it is enough to prove the case $|n| = 2$, but this case follows directly from the assumption. %Clearly, for any map $f$, if $w \in f(U \cap V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w \in f(U) \cap f(V)$. If $w \in f(U) \cap f(V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w = f(u) = f(v)$ for some $u \in U$ and $v \in V$, implying $u = v$ by injectivity of $f$. Hence $u \in U \cap V$, and $w \in f(U \cap V)$.
\end{proof}
%\vspace{-.2 cm}
\begin{lemma}\label{DistanceToIntersectionLemma}
Fix $k \geq 2$. Let $V_1, \ldots, V_k$ be subspaces of $\mathbb{R}^m$ and set $V = \cap_{i = 1}^k V_i$. For  $\mathbf{x} \in \mathbb{R}^m$, we have (where $\xi$ is given in Def.~\ref{SpecialSupportSet}):
\vspace{-.2 cm}
\begin{align}\label{DTILeq}
\text{\rm dist}(\mathbf{x}, V) \leq \frac{1}{\xi(\{V_i\}_{i = 1}^k)} \sum_{i=1}^k \text{\rm dist}(\mathbf{x}, V_i).
\end{align}
\end{lemma}
%\vspace{-.3 cm}
\begin{proof} 
Recall the orthogonal projection onto a subspace $V \subseteq \mathbb{R}^m$ is the mapping $\Pi_V: \mathbb{R}^m \to V$ that associates with each $\mathbf{x}$ its unique nearest point in $V$; i.e., $\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 = \text{\rm dist}(\mathbf{x}, V)$.
Next, observe:
\begin{align}\label{f}
\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 &\leq \|\mathbf{x} - \Pi_{V_k} \mathbf{x}\|_2 + \|\Pi_{V_k}  \mathbf{x} - \Pi_{V_k}\Pi_{V_{k-1}}\mathbf{x}\|_2 \nonumber \\
&\ \ \ + \cdots + \|\Pi_{V_k} \Pi_{V_{k-1}}\cdots \Pi_{V_1} \mathbf{x} - \Pi_V \mathbf{x}\|_2 \nonumber \\
&\leq \|\Pi_{V_k}\cdots\Pi_{V_{1}} \mathbf{x} - \Pi_V \mathbf{x}\|_2 + \sum_{\ell=1}^k \|\mathbf{x} - \Pi_{V_{\ell}} \mathbf{x}\|_2,
\end{align}
using the triangle inequality and that the spectral norm satisfies $\|\Pi_{V_{\ell}}\|_2 \leq 1$ for all $\ell$ (since $\Pi_{V_{\ell}}$ are orthogonal projections).

>>>>

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{B}lind source separation is a classical problem in signal processing \cite{sato1975method}.
In one common modern formulation, each of $N$ observed $n$-dimensional signals is a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown dictionary of size $m$, typically with $m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models).
Approximating solutions to this sparsity-constrained inverse problem have provided insight into the structure of many signal classes lacking domain-specific formal models (e.g., in vision \cite{wang2015sparse}).  In particular, it has been shown that response properties of simple-cell neurons in mammalian visual cortex emerge from optimizing a dictionary to represent small patches of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major advance in computational neuroscience. A curious aspect of this finding is that the latent waveforms (e.g., ``Gabor'' wavelets) estimated from data appear to be canonical \cite{donoho2001can};
i.e., they are found in learned dictionaries independent of algorithm or image training set.

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if all dictionaries consistent with training data do not differ appreciably in their ability to sparsely encode new samples. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain this uniqueness more broadly; in particular, a universal guarantee in the context of noise has yet to emerge.

@@

\begin{problem}[Sparse linear coding]\label{InverseProblem}
Find a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ so that $\|\mathbf{z}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \eta$ for all $i$.
\end{problem}

@@

%\vspace{-.15 cm}

@@

\begin{lemma}\label{spanIntersectionLemma}
If $f: U \to V$ is an injective function, then $f\left(\cap_{i=1}^\ell U_i \right) =  \cap_{i=1}^\ell f\left(U_i\right)$ for any $U_1, \ldots, U_\ell \subseteq U$.
%In particular, if for some $E \in 2^{[m]}$ the map $M \in \mathbb{R}^{n \times m}$ is injective on $\cup_{S \in \mathcal{H}} \text{span}\{e_i\}_{i \in S}$ then $\text{span}\{ M_{\cap_{S \in \mathcal{H}} S} \} = \cap_{S \in \mathcal{H}} \text{span}\{M_S\}$.
\end{lemma}
\begin{proof}
By induction it is enough to prove the case $|n| = 2$, but this case follows directly from the assumption. %Clearly, for any map $f$, if $w \in f(U \cap V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w \in f(U) \cap f(V)$. If $w \in f(U) \cap f(V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w = f(u) = f(v)$ for some $u \in U$ and $v \in V$, implying $u = v$ by injectivity of $f$. Hence $u \in U \cap V$, and $w \in f(U \cap V)$.
\end{proof}
%\vspace{-.2 cm}
\begin{lemma}\label{DistanceToIntersectionLemma}
Fix $k \geq 2$. Let $V_1, \ldots, V_k$ be subspaces of $\mathbb{R}^m$ and set $V = \cap_{i = 1}^k V_i$. For  $\mathbf{x} \in \mathbb{R}^m$, we have (where $\xi$ is given in Def.~\ref{SpecialSupportSet}):
%\vspace{-.2 cm}
\begin{align}\label{DTILeq}
\text{\rm dist}(\mathbf{x}, V) \leq \frac{1}{\xi(\{V_i\}_{i = 1}^k)} \sum_{i=1}^k \text{\rm dist}(\mathbf{x}, V_i).
\end{align}
\end{lemma}
%\vspace{-.3 cm}
\begin{proof} 
Recall the orthogonal projection onto a subspace $V \subseteq \mathbb{R}^m$ is the mapping $\Pi_V: \mathbb{R}^m \to V$ that associates with each $\mathbf{x}$ its unique nearest point in $V$; i.e., $\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 = \text{\rm dist}(\mathbf{x}, V)$.
Next, observe:
\begin{align}\label{f}
\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 &\leq \|\mathbf{x} - \Pi_{V_k} \mathbf{x}\|_2 + \|\Pi_{V_k}  \mathbf{x} - \Pi_{V_k}\Pi_{V_{k-1}}\mathbf{x}\|_2 \nonumber \\
&\ \ \ + \cdots + \|\Pi_{V_k} \Pi_{V_{k-1}}\cdots \Pi_{V_1} \mathbf{x} - \Pi_V \mathbf{x}\|_2 \nonumber \\
&\leq \|\Pi_{V_k}\cdots\Pi_{V_{1}} \mathbf{x} - \Pi_V \mathbf{x}\|_2 + \sum_{\ell=1}^k \|\mathbf{x} - \Pi_{V_{\ell}} \mathbf{x}\|_2,
\end{align}
using the triangle inequality and that the spectral norm satisfies $\|\Pi_{V_{\ell}}\|_2 \leq 1$ for all $\ell$ (since $\Pi_{V_{\ell}}$ are orthogonal projections).
